The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2012-05-14 smurf_paper.tex
Update Dempsey and Holland to 2012

The Submillimetre Common User Bolometer Array 2
\citep[\scuba,][]{holland2006} is a new instrument for the 15-m James
Clerk Maxwell Telescope (JCMT) on Mauna Kea, Hawai'i. The camera
simultaneously images the sky in two broad bands centered over 450 and
850\,\micron. The focal planes at each wavelength are populated with 4
rectangular subarrays, consisting of $40 \times 32$ bolometers each,
and subtend a nearly contiguous $4' \times 4'$ footprint on the sky.
Ultimately, the instrument will also be equipped with a Fourier
transform spectrometer \citep[FTS-2,][]{gom2010}, and a polarimeter
\cite[POL-2,][]{bastien2005}. This paper describes the Submillimetre
User Reduction Facility, SMURF, a software package for reducing the
imaging data. The reduction of FTS-2 and POL-2 data will be described
at a later date once these additional instruments have been
commissioned. The details of the instrument design, performance, and
its calibration are given in two companion papers: \citet{holland2011}
and \citet{dempsey2011}.

@@

While the details of how \scuba\ works is described in
\citet{holland2011}, and its calibration in \citet{dempsey2011}, we
summarize the basic concepts relative to map-making here.

@@

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing difficulties, considerable effort has gone into
dynamically applying position-dependent biases as well as an over-all
focal plane bias to make as many detectors useable as possible
\citep{holland2011}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation depending on
how the bias and amplification parameters were set up. To track these
responsivity changes, at the start of each observation, the current is
ramped through the bolometer heaters such that $dI/dP$ can be
measured. Note that the bolometers may have a slightly different
response to heater power due to conductance as compared to the
response to optical power, so additioanl efficiency corrections must
also be obtained. While not discussed further in this paper, details
may be found in \citet{dempsey2011}. We will refer to measurements of
$dI/dP$ across the focal plane as the
\emph{flatfield}. Fig.~\ref{fig:sensitivities} shows the flatfield for
each focal plane.

@@

The final conversion of power received by the detector into
astronomical flux units requires two more fundamental corrections: (i)
atmospheric extinction (due to atmospheric water vapour) must be
compensated for; and (ii) the corrected power is related to an
astronomical flux density through the effective area of the detector,
the spectral response of the bolometers (bandpass), the spectrum of
the source, and additional optical efficiency factors -- a \emph{flux
  conversion factor} (FCF). Extinction is measured, in most cases,
using the JCMT line-of-sight water vapour radiometer (WVM) which is
calibrated using skydips and the CSO 225\,GHz tipper. The FCF is
determined empirically using measurements of standard sources. For
further details on these calibration steps, see \citet{dempsey2011}.

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2011}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly from the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be estimated in the
common-mode estimation phase.

@@

The extinction correction is a multiplicative factor that is normally
derived using the JCMT water vapour radiometer (WVM), and is not
considered to be a free parameter in the solution. However, it is
applied as part of the iterative solution, rather than a
pre-processing step, since any small errors will be amplified by the
large low-frequency drifts in the raw bolometer time-series. For
example, if the bolometer drift is 1000 times greater than an
astronomical source of interest, a 1 per cent error in the flatfield
will produce stripes of order 10 times the astronomical signal
amplitude in the final map! If \model{EXT} is applied after the bulk
of the low-frequency noise has been removed (e.g., \model{COM,GAI}),
then there is little potential for such small errors to affect the
final map.  For details on how it is calculated see
\citet{dempsey2011}. Note that its numerical value is calculated only
once, and simply applied as a scale factor in the iterative solution.

>>>>

The Submillimetre Common User Bolometer Array 2
\citep[\scuba,][]{holland2006} is a new instrument for the 15-m James
Clerk Maxwell Telescope (JCMT) on Mauna Kea, Hawai'i. The camera
simultaneously images the sky in two broad bands centered over 450 and
850\,\micron. The focal planes at each wavelength are populated with 4
rectangular subarrays, consisting of $40 \times 32$ bolometers each,
and subtend a nearly contiguous $4' \times 4'$ footprint on the sky.
Ultimately, the instrument will also be equipped with a Fourier
transform spectrometer \citep[FTS-2,][]{gom2010}, and a polarimeter
\cite[POL-2,][]{bastien2005}. This paper describes the Submillimetre
User Reduction Facility, SMURF, a software package for reducing the
imaging data. The reduction of FTS-2 and POL-2 data will be described
at a later date once these additional instruments have been
commissioned. The details of the instrument design, performance, and
its calibration are given in two companion papers: \citet{holland2012}
and \citet{dempsey2012}.

@@

While the details of how \scuba\ works is described in
\citet{holland2012}, and its calibration in \citet{dempsey2012}, we
summarize the basic concepts relative to map-making here.

@@

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing difficulties, considerable effort has gone into
dynamically applying position-dependent biases as well as an over-all
focal plane bias to make as many detectors useable as possible
\citep{holland2012}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation depending on
how the bias and amplification parameters were set up. To track these
responsivity changes, at the start of each observation, the current is
ramped through the bolometer heaters such that $dI/dP$ can be
measured. Note that the bolometers may have a slightly different
response to heater power due to conductance as compared to the
response to optical power, so additioanl efficiency corrections must
also be obtained. While not discussed further in this paper, details
may be found in \citet{dempsey2012}. We will refer to measurements of
$dI/dP$ across the focal plane as the
\emph{flatfield}. Fig.~\ref{fig:sensitivities} shows the flatfield for
each focal plane.

@@

The final conversion of power received by the detector into
astronomical flux units requires two more fundamental corrections: (i)
atmospheric extinction (due to atmospheric water vapour) must be
compensated for; and (ii) the corrected power is related to an
astronomical flux density through the effective area of the detector,
the spectral response of the bolometers (bandpass), the spectrum of
the source, and additional optical efficiency factors -- a \emph{flux
  conversion factor} (FCF). Extinction is measured, in most cases,
using the JCMT line-of-sight water vapour radiometer (WVM) which is
calibrated using skydips and the CSO 225\,GHz tipper. The FCF is
determined empirically using measurements of standard sources. For
further details on these calibration steps, see \citet{dempsey2012}.

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2012}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly from the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be estimated in the
common-mode estimation phase.

@@

The extinction correction is a multiplicative factor that is normally
derived using the JCMT water vapour radiometer (WVM), and is not
considered to be a free parameter in the solution. However, it is
applied as part of the iterative solution, rather than a
pre-processing step, since any small errors will be amplified by the
large low-frequency drifts in the raw bolometer time-series. For
example, if the bolometer drift is 1000 times greater than an
astronomical source of interest, a 1 per cent error in the flatfield
will produce stripes of order 10 times the astronomical signal
amplitude in the final map! If \model{EXT} is applied after the bulk
of the low-frequency noise has been removed (e.g., \model{COM,GAI}),
then there is little potential for such small errors to affect the
final map.  For details on how it is calculated see
\citet{dempsey2012}. Note that its numerical value is calculated only
once, and simply applied as a scale factor in the iterative solution.
